---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Student Name"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---


## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1
getwd()

library(tidyverse)
library(agricolae)
library(lubridate)

NTL_LTER_Lake_ChemistryPhysics <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
         stringsAsFactors = TRUE)

#2
mytheme <- theme_bw(base_size = 13)+
  theme(plot.title = element_text(hjust = 0), 
        axis.text = element_text(color = "black"))

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Lake mean temperature in July does not change with depth across all lakes; regression coefficient = 0.
Ha: Lake mean temperature in July does change with depth across all lakes; regression coefficient ≠ 0.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_LTER_Lake_ChemistryPhysics$sampledate <- as.Date(NTL_LTER_Lake_ChemistryPhysics$sampledate, format = "%m/%d/%y")

NTL_LTER_Lake_ChemistryPhysics <- NTL_LTER_Lake_ChemistryPhysics %>% 
  mutate(month = format(sampledate, "%m"))

NTL_LTER_Lake_ChemistryPhysics$month <- 
  as.factor(NTL_LTER_Lake_ChemistryPhysics$month)

NTL_LTER_Lake_ChemistryPhysics_wrangled <- NTL_LTER_Lake_ChemistryPhysics %>% 
  filter(month == "07") %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit()

#5
NTL_LTER_temp_depth <- ggplot(NTL_LTER_Lake_ChemistryPhysics_wrangled, 
                              aes(x = depth, y = temperature_C))+
  geom_point(color = "blue")+
  ylim(0, 35)+
  geom_smooth(method = lm, color="black")+
  labs(title = "Temperature by Depth Across Lakes", x = "Depth (m)", 
       y = "Temperature (°C)")

print(NTL_LTER_temp_depth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure, if considering the line of best fit alone, suggests a negative association between the two variables; however, if you also consider the distribution of points, it appears that this trend is nonlinear. Overall, as depth increases, the temperature appears to decrease. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
linear_regression_depth_temp <- lm(data = NTL_LTER_Lake_ChemistryPhysics_wrangled, temperature_C ~ depth)
  summary(linear_regression_depth_temp)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: As depth increases by one meter there is an associated decrease in temperature of 1.95 degrees celsius. 73.9% of variation in temperature is explained by depth. The F-statistic is 2.75e4 with a degree of freedom of 9726. The p-value is 2.2e-16 which is well below the significance value of 0.001, suggesting that there is a highly significant relationship between temperature and depth and we can reject the null hypothesis. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp_AIC <- lm(data = NTL_LTER_Lake_ChemistryPhysics_wrangled, 
               temperature_C ~ year4 + daynum + depth)

step(temp_AIC)

#10
multiple_linear_temp <- lm(data = NTL_LTER_Lake_ChemistryPhysics_wrangled, 
                           temperature_C ~ daynum + depth + year4)

summary(multiple_linear_temp)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: AIC values increase when removing year4, daynum, and depth. This increase in AIC suggests that the removal of these variables made the model worse. In result, I included all three explanatory variables. This model explains a variance of 74.1% which is an improvement on the previous model which only uses depth to explain change in temperature. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
temperature_lakename_anova <- aov(data = NTL_LTER_Lake_ChemistryPhysics_wrangled, temperature_C ~ lakename)
summary(temperature_lakename_anova)

temperature_lakename_anova_lm <- lm(data = NTL_LTER_Lake_ChemistryPhysics_wrangled, temperature_C ~ lakename)
summary(temperature_lakename_anova_lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: A p-value of 2e-16 was achieved when running an anova test for temperature data among different lakes, meaning that there is significant difference in mean temperature among the different lakes. We do not know, however, which lakes differ in mean temperature value. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(NTL_LTER_Lake_ChemistryPhysics_wrangled, aes(x = depth, 
                                                    y = temperature_C, 
                                                    color = lakename))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm, se = FALSE)+
  ylim(0,35)+
  labs(title = "Temperature by Depth in Different Lakes", 
       x = "Depth (m)", 
       y = "Temperature (°C)", 
       color = "Lake Name")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(temperature_lakename_anova)

temp_lakename.groups <- HSD.test(temperature_lakename_anova, "lakename", group = TRUE)
temp_lakename.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul and Ward lakes have the same mean temperature as Peter Lake, statistically speaking. There is not a single lake which has a mean temperature which is statistically distinct from all other lakes evaluated with this model. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: To compare the means of Peter and Paul lakes we could run a two-sample t-test. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
NTL_LTER_Lake_ChemistryPhysics_Crampton.Ward <- NTL_LTER_Lake_ChemistryPhysics_wrangled %>% 
  filter(lakename =="Crampton Lake" | lakename == "Ward Lake")

twosample.t.test_Crampton.Ward <- t.test(NTL_LTER_Lake_ChemistryPhysics_Crampton.Ward$temperature_C ~ NTL_LTER_Lake_ChemistryPhysics_Crampton.Ward$lakename)
twosample.t.test_Crampton.Ward
```

>Answer: With a p-value of 0.265 there is not a statistically significant difference between the mean temperature values for Crampton and Ward lakes; we fail to reject the null hypothesis. This result matches the mean groupings generated in part 16 when running the HSD test. 
