---
title: "Assignment 5: Data Visualization"
author: "Kallie Davis"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()

library(tidyverse)
library(lubridate)
library(cowplot)

NTL_LTER_Lake_Chemistry_PeterPaul <-  read.csv("../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
NEON_NIWO_Litter_mass_trap <- read.csv("../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
NTL_LTER_Lake_Chemistry_PeterPaul$sampledate <- as.Date(NTL_LTER_Lake_Chemistry_PeterPaul$sampledate, format = "%Y-%m-%d") 
NEON_NIWO_Litter_mass_trap$collectDate <- as.Date(NEON_NIWO_Litter_mass_trap$collectDate, format = "%Y-%m-%d")
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_bw(base_size = 15) +
  theme(axis.text = element_text(color = "maroon"))

theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4
p <- ggplot(subset(NTL_LTER_Lake_Chemistry_PeterPaul, lakename == "Paul Lake" | lakename == "Peter Lake"),
          aes(x = po4, y = tp_ug))+
        geom_point(aes(color = lakename))+
          labs(x = "Phosphate", y = "Total Phosphorus")+
          geom_smooth(method = lm, color="black")+
          xlim(1, 45)

print(p)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5
#temperature plot
NTL_LTER_tp_temp <- ggplot(NTL_LTER_Lake_Chemistry_PeterPaul)+
  geom_boxplot(aes(x = factor(month), y = temperature_C, color = lakename))

print(NTL_LTER_temp)

#total phosphorus plot
NTL_LTER_tp_ug <- ggplot(NTL_LTER_Lake_Chemistry_PeterPaul)+
  geom_boxplot(aes(x = factor(month), y = tp_ug, color = lakename))

print(NTL_LTER_tp_ug)

#total nitrogen plot
NTL_LTER_tn_ug <- ggplot(NTL_LTER_Lake_Chemistry_PeterPaul)+
  geom_boxplot(aes(x = factor(month), y = tn_ug, color = lakename))

print(NTL_LTER_tn_ug)

```

```{r, fig.height=20, fig.width=10}

#combined plot
NTL_LTER_combined <- plot_grid(NTL_LTER_temp + theme(legend.position = "none"),
                               NTL_LTER_tp_ug + theme(legend.position = "none"),
                               NTL_LTER_tn_ug+ theme(legend.position = "none"), 
                               nrow = 3, ncol = 1, 
                               align = "v")


print(NTL_LTER_combined)
```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: Total nitrogen and phosphorus have relatively consistent median values across seasons. Median total phosphorus and nitrogen values are greater in Peter lake.  

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6
Needles_dryMass_CollectDate <- ggplot(subset(NEON_NIWO_Litter_mass_trap, functionalGroup == "Needles"))+
        geom_point(aes(x = collectDate, y = dryMass,color = nlcdClass))+
          labs(x = "Collect Date", y = "Dry Mass")

print(Needles_dryMass_CollectDate)

#7
Needles_dryMass_CollectDate_faceted <- ggplot(subset(NEON_NIWO_Litter_mass_trap, functionalGroup == "Needles"))+
        geom_point(aes(x = collectDate, y = dryMass))+
          labs(x = "Collect Date", y = "Dry Mass")+
     facet_wrap(vars(nlcdClass), nrow = 3)

print(Needles_dryMass_CollectDate_faceted)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: