---
title: "Assignment 2: Coding Basics"
author: "Kallie Davis"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. The sequence (seq) function is being used to generate a sequence of values
# going from 1 to 100, increasing by fours.
sequence_1 <- seq(1, 100, 4)

#2. Calculated the mean and median using the respective functions. 
sequence_1_mean <- mean(sequence_1)
sequence_1_median <- median(sequence_1)
sequence_1_mean
sequence_1_median

#3. I created a logical statement to see if the sequence median and mean values
# were the same. 
sequence_1_mean == sequence_1_median

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
student_names <- c("Amelia", "Sophia", "Charlie", "Sam")       #character
test_scores <- c(64,78,34,97)                                  #numeric
passing_grade <- c(FALSE, TRUE, FALSE, TRUE)                    #logical

student_test_grades.df <- data.frame(student_names,test_scores,passing_grade)
names(student_test_grades.df) <- c(" Student Name","Test Score","Passing Grade")
head(student_test_grades.df)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame is capable of containing different data classes; a matrix can only contain a single data class.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}

calc_passing_grade <- function(x) {
  ifelse(x<50, FALSE, TRUE)
}

pass_or_fail <- calc_passing_grade(student_test_grades.df$`Test Score`); pass_or_fail
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The `ifelse` option worked. The `if` and `else` function will only work with one input value from a vector at a time; the ifelse function will evaluate all values in a vector at once.
